Astrophysical calibration of gravitational-wave detectors

### 2015-04-23 docs/cosmic_cal.tex
Some very minor changes - mainly adding some references

\begin{abstract}
  We present an investigation into the potential of calibrating
  gravitational wave detector outputs through the use of standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  obtained and combined with current knowledge of the cosmologic
  parameters.  This will yield the true luminosity distance and allow
  the direct comparison with the estimated value.  Discrepancies can
  then be used to correct the original calibration.~\cm{Update at end}
\end{abstract}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect such waves emitted from $O(10s)$ of compact binary
coalescences.  One such class of these cataclysmic events, the inspiral and
merger of \ac{BNS} systems will be detected out to a maximum range of $\approx
450$ Mpc. Assuming the current best estimates for the cosmological parameters,
this is equivelent to a redshift $z\approx 0.1$. As noted by~\cite{} \ac{CBC}
systems can be used as cosmiological distance markers, otherwise known as
"standard sirens" (analogous to the \ac{EM} standard candles). Standard sirens
allow ...

@@

% talk about the calibration idea
If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it will
be possible to identify the host galaxy of the \ac{BNS} source.  With this
information a spectroscopic redshift can be very accurately obtained. With
knowledge of the redshift, using current best estimates for the Hubble constant
and other cosmological parameters, the true luminosity distance can be
estimated to $\sim 1\%$ accuracy  The \ac{GW} measurement acts as a standard
siren also giving us a direct measurement of the luminosity distance to the source.
The accuracy of such a measurement depends on a number of factors including the
accuracy with which the \ac{GW} detector has been calibrated.  Hence comparison
with the distance estimate from the \ac{sGRB} we can re-calibrate (or validate)
the existing experimentally obtained calibration.

@@

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see~\cite{Vitale} and references therein. For advanced LIGO and Virgo,
at the frequency range used in this project, the error in the calibration using
the 'hardware injection method' is roughly $10\%$~\cite{Vitale}. This is a
benchmark set for the error estimation using the proposed new method in this
project.~\cm{Clean this up and expand.}

% Describe our model for the calibration
In this work the calibration correction factor used throughout is simply a scale
factor, $s$. This effects the data set collected by a simple multiplication:
%
\begin{equation}\label{eq:scaledata}
  h(t) = s\times h_{\mathrm{m}}(t)
\end{equation}
%
where $h_{\mathrm{m}}(t)$ is the measured detector strain, and
$h(t)$ is the exact perfectly calibrated strain. A description of the
collected data is given in section~\ref{sec:cbc}. The scale factor being
parameterised, $s$, represents all known sources of error in the system. This
is not entirely realistic as $s$ is known to be a function of frequency.
Testing a more realistic calibration error is work to be done in the
future.~\cm{Need to clean this up and expand the discussion}

@@

% discuss the inclination angle issue
A fortunate consequence of a joint \ac{GW}--\ac{sGRB} observation will be the
fact that in order for such an event to be observed, the \ac{BNS} system must
have had its orbital angular momentum vector pointing towards (or away) from
the detector.  The actual inclination angle of the system, defined as the angle
between the orbital angular momentum vector and the line of sight, must be $<$
half of the \ac{sGRB} beaming angle.  This prerequisite property limits us to
systems that are approximately "face-on" and therefore biases us to higher
\ac{SNR} signals.  However, as discussed earlier, the property of beaming
severely impacts the probable rate of such joint observations.

@@

>>>>

\begin{abstract}
  We present an investigation into the potential of calibrating \MP{change to: ``assessing the validity of the calibration''?}
  gravitational wave detector outputs through the use of standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  obtained and combined with current knowledge of the cosmologic
  parameters.  This will yield the true luminosity distance and allow
  the direct comparison with the estimated value.  Discrepancies can
  then be used to correct the original calibration.~\cm{Update at end}
\end{abstract}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect waves emitted from $O(10s)$ of compact binary
coalescences.  One such class of these cataclysmic events, the inspiral and
merger of \ac{BNS} systems will be detected out to a maximum range of $\approx
450$ Mpc. Assuming the current best estimates for the cosmological parameters,
this is equivelent to a redshift $z\approx 0.1$. As noted by~\cite{} \ac{CBC}
systems can be used as cosmiological distance markers, otherwise known as
``standard sirens'' (analogous to the \ac{EM} standard candles). Standard sirens
allow ...

@@

% talk about the calibration idea
If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it will \MP{may?}
be possible to identify the host galaxy of the \ac{BNS} source.  With this
information a spectroscopic redshift can be very accurately obtained. With
knowledge of the redshift, using current best estimates for the Hubble constant
and other cosmological parameters, the true luminosity distance can be
estimated to $\sim 1\%$ accuracy.  The \ac{GW} measurement acts as a standard
siren also giving us a direct measurement of the luminosity distance to the source.
The accuracy of such a measurement depends on a number of factors including the
accuracy with which the \ac{GW} detector has been calibrated.  Hence comparison
with the distance estimate from the \ac{sGRB} we can re-calibrate (or validate)
the existing experimentally obtained calibration.

@@

\MP{In this paper I think we stick to just talking about a single overall amplitude calibration scale.}

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see~\cite{Vitale:2012} and references therein. For advanced LIGO and Virgo,
at the frequency range used in this project, the error in the calibration using
the `hardware injection method' is roughly $10\%$~\cite{Vitale:2012}. This is a
benchmark set for the error estimation using the proposed new method in this
project.~\cm{Clean this up and expand.}

% Describe our model for the calibration
In this work the calibration correction factor used throughout is simply a scale
factor, $s$. This effects the data set collected by a simple multiplication:
%
\begin{equation}\label{eq:scaledata}
  h(t) = s\times h_{\mathrm{m}}(t)
\end{equation}
%
where $h_{\mathrm{m}}(t)$ is the measured detector strain, and
$h(t)$ is the exact perfectly calibrated strain. A description of the
collected data is given in Section~\ref{sec:cbc}. The scale factor being
parameterised, $s$, represents all known sources of error in the system. This
is not entirely realistic as $s$ is known to be a function of frequency.
Testing a more realistic calibration error is work to be done in the
future.~\cm{Need to clean this up and expand the discussion}

@@

% discuss the inclination angle issue
A fortunate consequence of a joint \ac{GW}--\ac{sGRB} observation will be the
fact that in order for such an event to be observed, the \ac{BNS} system must
have had its orbital angular momentum vector pointing towards (or away) from
the detector.  The actual inclination angle of the system, defined as the angle
between the orbital angular momentum vector and the line of sight, must be $<$
half of the \ac{sGRB} beaming angle.  This prerequisite property limits us to
systems that are approximately ``face-on'' and therefore biases us to higher
\ac{SNR} signals.  However, as discussed earlier, the property of beaming
severely impacts the probable rate of such joint observations.

@@

\MP{We emphasise that our work has assumed that the PSD used in the likelihood function
for estimating the calibration scale factor is an accurate representation of the
PSD at the time of the observed CBC signal. In reality the PSD may be estimated from
a period close to, but not overlapping, the signal time and therefore may be slightly
different (see e.g. \cite{2013PhRvD..88h4044L}). Our result would therefore be completely correlated
with any uncertainty in the PSD estimate, but would still offer an upper limit
on the calibration uncertainty.}

\MP{The BayesLine algorithm \cite{2015PhRvD..91h4034L} may provide a natural
way to perform this analysis in particular when accounting for a frequency
dependent calibration.}
